Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

### 2024-10-31 ms/ms.Rmd
ms: results - depthByClass description

We obtained depth measurements for 372 cells (XX % of mapped peatland) at Skrimfjella and 1889 cells (XX % of mapped peatland) at Ørskogfjellet.
The mean depths of these \unit{10\,m} cells differed by only \unit{6\,cm} between Skrimfjella and Ørskogfjellet (Table \ref{tab:depthsByClass}).

>>>>

We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 % of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 % of mapped peatland).
Roughly 80 % of these \unit{10\,m} cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 % versus 27 % of cells), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet:  \unit{119\,cm} and \unit{126\,cm}, respectively.

### 2024-10-31 ms/ms.tex
ms: right-censored probes
ms: results - depthByClass description

Field work at Skrimfjella was conducted in August 2020.
We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
We used a metal probe pushed downward until resistance indicated the base of the peat column.
Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
We judged whether the soil was a peat soil based on its density, texture, and color.
Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.

@@

We obtained depth measurements for 372 cells (XX \% of mapped peatland) at Skrimfjella and 1889 cells (XX \% of mapped peatland) at Ørskogfjellet.
The mean depths of these \unit{10\,m} cells differed by only \unit{6\,cm} between Skrimfjella and Ørskogfjellet (Table \ref{tab:depthsByClass}).

\begin{table}[tbp]
\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
\begin{tabular}{llllllll}
\hline
    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
    &                            & 372      & 100     & 119        & 1889       & 100     & 125        \\
AR5 & Fully cultivated           &          &         &            & 21         & 1       & 30         \\
    & Forest                     & 52       & 14      & 65         & 272        & 14      & 71         \\
    & Open firm ground           & 19       & 5       & 98         & 134        & 7       & 39         \\
    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
DMK & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
    & shallow (\textless 100 cm) & 6        & 2       & 50         & 839        & 44      & 82         \\
    & unknown                    & 272      & 73      & 127        & 391        & 21      & 59         \\ \hline
\end{tabular}
\label{tab:depthsByClass}
\end{table}

>>>>

Field work at Skrimfjella was conducted in August 2020.
We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
We used a metal probe pushed downward until resistance indicated the base of the peat column.
Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
A single probe measurement was right-censored because the peat column was deeper than the probe.
For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
We judged whether the soil was a peat soil based on its density, texture, and color.
Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.

@@

We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 \% of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 \% of mapped peatland).
Roughly 80 \% of these \unit{10\,m} cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 \% versus 27 \% of cells), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet: \unit{119\,cm} and \unit{126\,cm}, respectively.

\begin{table}[tbp]
\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
\begin{tabular}{llllllll}
\hline
    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
    &                            & 372      & 100     & 119        & 1878       & 100     & 126        \\
AR5 & Agricultural               &          &         &            & 21         & 1       & 30         \\
    & Forest                     & 52       & 14      & 65         & 272        & 15      & 71         \\
    & Open upland                & 19       & 5       & 98         & 134        & 7       & 39         \\
    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
DMK & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
    & shallow (\textless 100 cm) & 6        & 2       & 50         & 838        & 45      & 82         \\
    & unknown                    & 272      & 73      & 127        & 381        & 20      & 60         \\ \hline
\end{tabular}
\label{tab:depthsByClass}
\end{table}
